Parsing Postfix log files

### 2009-03-17 doc/logparser-implementation.tex
Added a diagram of the database schema.  It took a long time to find a tool to draw it with, longer to draw it, and even longer still to make it fit on the page properly.
Checked all references in the implementation chapter.

Finished the flow chart introduction.

Some minor changes and corrections.

XXX UPDATE REFERENCES WHEN ARCHITECTURE AND IMPLEMENTATION ARE MORE FLESHED
OUT\@.  XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE
CORRECT CHAPTER\@.

The parser deals with all the complications of parsing, the eccentricities
and oddities of Postfix log files, presenting the resulting data in a
normalised, simple to use representation.  The parser's task is to follow
the journey each mail takes through Postfix, combining the data extracted
by rules into a coherent whole, saving it in a useful and consistent form,
and performing housekeeping duties.

@@

XXX NEEDS AN INTRO\@.

A picture is said to be worth a thousand words: the flow chart in
\figureref{flow chart image} shows the most common paths a connection or
mail can take through \parsername{}; the parsing complications described in
\sectionref{complications} are excluded for the sake of clarity.

@@

    \item If one or more mails were accepted there will be more log lines
        for those mails later, see \sectionref{mail delivery}.

@@

\label{mail delivery}

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{10};
        \sectionref{tracking re-injected mail} discusses this in detail.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{12}.  This is the last log line
        for this particular mail (though it may be indirectly referred to
        if it was re-injected).  If it is neither parent nor child of
        re-injection the data is cleaned up and entered in the database,
        then deleted from the state tables.  The handling of re-injected
        mails is described in \sectionref{tracking re-injected mail}.

@@

The database schema can be conceptually divided in two: the rules used to
recognise log lines, and the data saved from the parsing of log files.
Rules have the fields required to recognise log lines, extract data to be
saved, and specify the action to be invoked; they also have several fields
that aid the user in understanding the meaning of the log lines recognised
by each rule.  The rules are described in detail in \sectionref{rules in
implementation}, but the fields are covered with the rest of the database
schema in this section.

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection where there was a rejection; the
individual fields will be described in \sectionref{connections table}.
There will be at least one entry in the results table for each entry in the
connections table; the result table's fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \sectionref{Diagram of the Database Schema} before the in-depth
descriptions of each table.

@@

A clear, comprehensible schema is essential when using the extracted data;
it is more important when using the data than when storing it, because
storing the data is a write-once operation, whereas utilising the data
requires frequent searching, sorting, and manipulation of the data to
produce customised reports or statistics.

@@

\subsection{Diagram of the Database Schema}

\label{Diagram of the Database Schema}

XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.

@@

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is covered here with the other tables in
the database.  Rules are created by the user, not the parser, and will not
be modified by the parser (except for the hits and hits\_total fields).
Rules classify the individual log lines, extracting data to be saved in the
connections and results tables, and specifying the action to take for that
log line.

@@

    \item [regex] The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of data; and makes each regex largely self-documenting.  For
        efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file; otherwise every
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown, as shown in \sectionref{Caching
        compiled regexes}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the value on the right hand
        side.\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be included in data.  This feature is intended for use with
        small amounts of data, so dealing with escape sequences was deemed
        unnecessary.}  Any field in the connections table can be set in
        this way.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing (described in \sectionref{Re-injected mails}).
        In the latter case the message-id will be unchanged from the
        original bounce notification, and so even if it happens to be in
        the correct format (e.g.\ if it was generated by Postfix on this or
        another server) it will not match the queueid of the mail.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but the aborted delivery attempts described in
this section only occur four times in the \numberOFlogFILES{} log files
used for testing.  The symptoms are the same as in \sectionref{aborted
delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
line; there does not appear to be anything in the log file to explain why
there are no further log lines.  The only way to detect these mails is to
periodically scan all mails in the state tables, deleting any mails
displaying the following characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop\glsadd{mail
loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
Postfix has had a chance to try to deliver it.  In this case the recipient
address will not have been logged, so a dummy recipient address needs to be
added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  This
complication has been observed in XXX/\numberOFlogFILES{} log files, but
typically when it does occur there were many instances of it, closely
grouped.

>>>>

The parser deals with all the complications of parsing, the eccentricities
and oddities of Postfix log files, presenting the resulting data in a
normalised, simple to use representation.  The parser's task is to follow
the journey each mail takes through Postfix, combining the data captured by
rules into a coherent whole, saving it in a useful and consistent form, and
performing housekeeping duties.

@@

A picture is said to be worth a thousand words: the flow chart in
\figureref{flow chart image} shows the most common paths a connection or
mail can take through \parsername{}; the parsing complications described in
\sectionref{complications} are excluded for the sake of clarity.  The flow
chart is intended to be a graphical overview of how a mail progresses
through both Postfix and \parsername{}, providing an overall context into
which the detailed descriptions in the remainder of this chapter will fit,
in particular the actions (\sectionref{actions in implementation}) and
complications (\sectionref{complications}).

@@

    \item If one or more mails were accepted there will be more log lines
        for those mails.

@@

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{10}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator \flowchart{COMMIT}{12}.  This is the last log line
        for this particular mail (though it may be indirectly referred to
        if it was re-injected).  If it is neither parent nor child of
        re-injection the data is cleaned up and entered in the database,
        then deleted from the state tables.

@@

The database schema can be conceptually divided in two: the rules used to
recognise log lines, and the data saved from the parsing of log files.
Each rule has a regex to recognise log lines and capture data from them,
and specifies the action to be invoked when a log line is recognised; they
also have several fields that aid the user in understanding the meaning of
the log lines recognised by each rule.  The rules are described in detail
in \sectionref{rules in implementation}, but the rules table is documented
in \sectionref{rules table} with the rest of the database schema.

The data saved from parsing the log files is also divided into two tables:
connections and results.  The connections table contains an entry for every
mail accepted and every connection where there was a rejection; the
individual fields will be described in \sectionref{connections table}.
There will be at least one entry in the results table for each entry in the
connections table; the result table's fields will be covered in detail in
\sectionref{results table}.  A diagram of the database schema is provided
in \figureref{Diagram of the Database Schema picture}, alongside the
in-depth descriptions of each table.

@@

A clear, comprehensible schema is essential when using the data extracted
from log lines; it is more important when using the data than when storing
it, because storing the data is a write-once operation, whereas utilising
the data requires frequent searching, sorting, and manipulation of the data
to produce customised reports or statistics.

@@

@@

\newlength{\belowcaptionskipORIG}
\setlength{\belowcaptionskipORIG}{\belowcaptionskip}
\setlength{\belowcaptionskip}{10pt}
\showgraph{database-schema}{Diagram of the Database Schema}{Diagram
of the Database Schema picture}
\setlength{\belowcaptionskip}{\belowcaptionskipORIG}

Rules are discussed in detail in \sectionref{rules in implementation}, but
the structure of the rules table is documented here along with the other
tables in the database.  Rules are created by the user, not the parser, and
will not be modified by the parser (except for the hits and hits\_total
fields).  Rules classify the individual log lines, capturing data to be
saved in the connections and results tables, and specifying the action to
take for that log line.

@@

    \item [regex] The regex to recognise log lines with.  The regex will
        have several keywords expanded when the rules are loaded: this
        simplifies reading and writing rules; avoids needless repetition of
        complex regex components; allows the components to be corrected
        and/or improved in one location; enables automatic extraction and
        saving of captured data; and makes each regex largely
        self-documenting.  For efficiency the keywords are expanded and
        every rule's regex is compiled before attempting to parse the log
        file; otherwise every regex would be recompiled each time it was
        used, resulting in a large, data dependent slowdown, as shown in
        \sectionref{Caching compiled regexes}.

    \item [connection\_data] Sometimes rules need to save data that is not
        present in the log line: e.g.\ setting \texttt{client\_ip} when a
        mail is being delivered to another server.  The format is:
        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
        i.e.\ semi-colon or comma separated assignment statements, with the
        variable name on the left and the value on the right hand side.
        Any field in the connections table can be set in this way.  Commas
        and semi-colons cannot be escaped and thus cannot be included in
        data.  This feature is intended for use with small amounts of data,
        so dealing with escape sequences was deemed unnecessary.

@@

    \item The queueid in the message-id must be the same as the queueid of
        the mail: this is what distinguishes bounce notifications generated
        locally from bounce notifications which are being re-injected as a
        result of aliasing.  In the latter case the message-id will be
        unchanged from the original bounce notification, and so even if it
        happens to be in the correct format (e.g.\ if it was generated by
        Postfix on this or another server) it will not match the queueid of
        the mail.

@@

The aborted delivery attempts described in \sectionref{aborted delivery
attempts} occur frequently, but the aborted delivery attempts described in
this section only occur four times in the \numberOFlogFILES{} log files
used for testing.  The symptoms are the same as in \sectionref{aborted
delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
line; there is nothing in the log file to explain why there are no further
log lines.  The only way to detect these mails is to periodically scan all
mails in the state tables, deleting any mails displaying the following
characteristics:

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop\glsadd{mail
loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
Postfix has had a chance to try to deliver it.  In this case the recipient
address will not have been logged, so a dummy recipient address needs to be
added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  Typically
when this complication occurs there are many instances of it, closely
grouped.
